Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-03-10 curvature.tex
dropped k-tubes

To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
A na\"ive approach would require $\OhOf{n^3}$ time---$OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves with a one-to-one correspondence to SPR neighbors, as described in the next paragraph.
Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
We note that the recursive formula of Song \cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.

@@

\section{$k$-tubes}

The curvature with respect to two topologies, $T_1$ and $T_2$, is primarily determined by a subset of the SPR graph in the localized neighborhood of these topologies.
We first require some notation to define these subsets.
Let $\TT$ be the full set of topologies with a given number of leaves.
Denote the set of trees with SPR distance at most $k$ from a given tree $T$ by $N_k(T)$.
These trees are the $k$-neighbors of $T$.
That is, $$ N_k(T) = \set{t \in \TT \mid \dspr{T, t} \le k} .$$

We now define the $k$-tube of two trees $T_1$ and $T_2$ to be the union of their $k$-neighbors and \emph{every tree on a shortest path between a $k$-neighbor of $T_1$ and a $k$-neighbor of $T_2$}.
More formally, the $k$-tube for $T_1$ and $T_2$ is:
\[ \set{t \in \TT \mid \dspr{t, t_1} + \dspr{t, t_2} = \dspr{t_1, t_2},\ \exists_{t_1 \in N_k(T_1)},\ \exists_{t_2 \in N_k(T_2)} } .\]
We call $T_1$ and $T_2$ the centers of the $k$-tube.
The distance between $k$-neighbors of different central trees is bounded by the maximum distance from each $k$-neighbor to its respective center and the maximum distance between the centers:

\begin{lemma}
	\label{lem:neighbor_distance}
	$\dspr{t_1, t_2} \le \dspr{T_1, T_2} + 2k,\ \forall_{t_1 \in N_k(T_1)},\ \forall_{t_2 \in N_k(T_2)}.$
\end{lemma}

To efficiently compute a $k$-tube, we first bound the maximum distance from members of the $k$-tube to the closest central tree.
\begin{lemma}
	\label{lem:k_tube_distance}
	Any tree $t$ of the $k$-tube of two trees $T_1$ and $T_2$ satisfies:
	\[ \min\left({\dspr{t, T_1}, \dspr{t, T_2}}\right) \le \floor{\dspr{T_1, T_2}/2} + 2k .\]
\end{lemma}
\begin{proof}
Without loss of generality, assume that $t$ is on a shortest path between two trees $t_1 \in N_k(T_1)$ and $t_2 \in N_k(T_2)$, and is either closer to $t_1$ than $t_2$ or equidistant from both.
By Lemma~\ref{lem:neighbor_distance} there is a path of length at most $\dspr{T_1, T_2} + 2k$ between $t_1$ and $t_2$.
Then $\dspr{t, t_1} \le \floor{(\dspr{T_1, T_2} + 2k) / 2}$.

Along with the fact that $\dspr{T_1, t_1} \le k$, we have that
\begin{align*}
\dspr{t, T_1} &\le \dspr{t, t_1} + \dspr{t_1, T_1} \\
&\le \floor{(\dspr{T_1, T_2} + 2k) / 2} + k \\
&\le \floor{\dspr{T_1, T_2} / 2} + 2k.
\end{align*}

\end{proof}
We can thus restrict our search for members of the $k$-tube to topologies of distance at most $\floor{\dspr{T_1, T_2}/2} + 2k$ from the central trees.
We call this set of trees $k$-tube candidates.
Note that there are $\OhOf{n^{2(\floor{\dspr{T_1, T_2}/2} + 2k)}}$ $k$-tube candidates.
We identify these candidates by iterating over increasingly large $k$-neighbourhoods of $T_1$ and $T_2$, and removing duplicate topologies through hashing.

In the second part of our procedure for computing a $k$-tube, we identify $k$-tube candidates that are not on the shortest path between two $k$-neighbors of the central trees.
We first compute shortest path lengths from each member of $N_k(T_1)$ to each $k$-tube candidate using breadth-first search (BFS).
For each BFS traversal tree, we backtrack from members of $N_k(T_2)$, identifying topologies in the $k$-tube.
We can use unique tree indices here, rather than hashing, to avoid duplicates.
The size of $N_k(T_1)$ is $\OhOf{n^{2k}}$, so this takes
$\OhOf{ n^{2k} M}$ time, where $M$ is the number of edges between $k$-tube candidates, that is, at most the square of the number of $k$-tube candidates.

This procedure identifies exactly the trees in the $k$-tube, so we have the following lemma:
\begin{lemma}
	\label{lem:k_tube_correctness}
	This procedure correctly identifies the $k$-tube.
\end{lemma}
\begin{proof}
	First, suppose that this procedure identifies a tree $x$ that is not in the $k$-tube.
	Assume, without loss of generality, that the BFS tree $B$ from tree $t_1 \in N_k(T_1)$ transitioned from $x$ to a $k$-tube tree $t$.
	Moreover, there must be a path in $B$ from $t$ to a tree $t_2 \in N_k(T_2)$ such that $t$ is on a shortest path from $t_1$ to $t_2$.
	Then, by the properties of BFS, $x$ is on this shortest path and belongs in the $k$-tube.

	Now, suppose that this procedure does not identify a tree $x$ that belongs in the $k$-tube.
	$x$ must be on a shortest path between some trees $t_1 \in N_k(T_1)$ and $t_2 \in N_k(T_2)$.
	Then, by the properties of BFS, $x$ would have been on the path from $t_1$ to $t_2$ in the $t_1$ BFS tree.
\end{proof}

>>>>

To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
A na\"ive approach would require $\OhOf{n^3}$ time---$\OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves with a one-to-one correspondence to SPR neighbors, as described in the next paragraph.
Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
We note that the recursive formula of Song \cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.

@@
